---
title: "cognition manuscript analyses"
author: "Alex Boren"
date: "2026-07-02"
output: html_document
---

This file contains the code for the analyses described in manuscript. 

Install packages/load libraries
```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(ggtext) 
library(scales) 
library(likert)
library(dunn.test) 
library(car) 
library(psych)
library(nnet)
library(MASS) 
library(vegan) 
library(purrr)
library(gtsummary)
library(effectsize)
library(patchwork)
library(pwr)
```

# Quantitative Analysis

Load quantitative data
```{r}
data_quant <- read.csv("data_main/data_combined.csv",
                       header=TRUE,
                       stringsAsFactors=TRUE, 
                       sep=",",
                       na.strings="NA",
                       dec=".",
                       strip.white=TRUE)
```



## PANAS scale analyses
```{r}
#Convert PANAS scale responses to numeric values
panas_conversion <- c(
  "Very slightly or not at all" = 1, 
  "A little" = 2, 
  "Moderately" = 3, 
  "Quite a bit" = 4, 
  "Extremely" = 5)

# Convert columns e_1 to e_10 to numeric
data_quant <- data_quant %>%
  mutate(across(e_1:e_10, ~ panas_conversion[as.character(.x)] %>% as.numeric()))

#Descriptive statistics (mean and standard deviation for each PANAS item per condition)
panas_summary <- data_quant %>%
  group_by(message) %>%
  summarise(across(e_1:e_10,
                   list(mean = ~mean(.x, na.rm = TRUE), 
                        sd   = ~sd(.x, na.rm = TRUE))))
panas_summary
```
Column names and corresponding emotions: 
e_1 inspired 
e_2 Afraid 
e_3 Alert
e_4 Upset 
e_5 Excited
e_6 Nervous 
e_7 Enthusiastic 
e_8 Scared 
e_9 Determined 
e_10 Distressed

Positive words: e_1 inspired, e_3 Alert, e_5 Excited, e_7 Enthusiastic, e_9 Determined 
Negative words: e_2 Afraid, e_4 Upset, e_6 Nervous, e_8 Scared, e_10 Distressed


Prepare PANAS data for ANOVA and MANOVA
```{r}
#create subsetted data that only includes study2 responses (only study2 included the PANAS question)
data_quant_2 <- data_quant %>%
  filter(type == "final")

#calculating means of positive and negative scores respectively 
data_summary <- data_quant_2 %>% 
  mutate(
    positive_mean = rowMeans(select(., e_1, e_3, e_5, e_7, e_9), na.rm = TRUE),
    negative_mean = rowMeans(select(., e_2, e_4, e_6, e_8, e_10), na.rm = TRUE))

#z-score means
data_summary <- data_summary %>%
  mutate(
    positive_mean_z = as.numeric(scale(positive_mean)),
    negative_mean_z = as.numeric(scale(negative_mean)))
```

Run univariate ANOVAs testing effect of conservation message * cognition message * species on mean positive and negative PANAS scores, respectively
```{r}
#First, need to create binary factors from message condition
data_summary <- data_summary %>%
  mutate(
    cognition    = factor(ifelse(message %in% c("Cognition", "Combined"), "yes", "no"),
                          levels = c("no", "yes")),
    conservation = factor(ifelse(message %in% c("Conservation", "Combined"), "yes", "no"),
                          levels = c("no", "yes")))

table(data_summary$message, data_summary$cognition, data_summary$conservation) #verify each message levels maps to one cell


###Positive Affect ANOVA
#create linear model
m_pos <- lm(positive_mean_z ~ cognition * conservation * species, data = data_summary)

#Run type III ANOVA, calculate effect size (partial eta square), and create results table
options(contrasts = c("contr.sum", "contr.poly"))

anova_results_pos <- Anova(m_pos, type = 3)
anova_df_pos <- as.data.frame(anova_results_pos)
anova_df_pos$Source <- rownames(anova_df_pos)
anova_df_pos$MS <- anova_df_pos$`Sum Sq` / anova_df_pos$Df

eta_df_pos <- as.data.frame(eta_squared(anova_results_pos, partial = TRUE))
eta_df_pos$Source <- eta_df_pos$Parameter

final_df_pos <- dplyr::left_join(anova_df_pos, eta_df_pos, by = "Source")
final_df_pos <- final_df_pos[, c("Source", "Sum Sq", "Df", "MS", "F value", "Pr(>F)", "Eta2_partial")]
final_df_pos <- final_df_pos %>%
  mutate(across(where(is.numeric), ~ round(., 3)))
final_df_pos[is.na(final_df_pos)] <- ""

print(final_df_pos)

###Negative Affect ANOVA
#create linear model
m_neg <- lm(negative_mean_z ~ cognition * conservation * species, data = data_summary)

#Run type III ANOVA, calculate effect size (partial eta square), and create results table
anova_results_neg <- Anova(m_neg, type = 3)
anova_results_neg
anova_df_neg <- as.data.frame(anova_results_neg)
anova_df_neg$Source <- rownames(anova_df_neg)
anova_df_neg$MS <- anova_df_neg$`Sum Sq` / anova_df_neg$Df

eta_df_neg <- as.data.frame(eta_squared(anova_results_neg, partial = TRUE))
eta_df_neg$Source <- eta_df_neg$Parameter

final_df_neg <- dplyr::left_join(anova_df_neg, eta_df_neg, by = "Source")
final_df_neg <- final_df_neg[, c("Source", "Sum Sq", "Df", "MS", "F value", "Pr(>F)", "Eta2_partial")]
final_df_neg <- final_df_neg %>%
  mutate(across(where(is.numeric), ~ round(., 3)))
final_df_neg[is.na(final_df_neg)] <- ""

print(final_df_neg)
```

Run MANOVA on mean positive and negative PANAS scores compositely
```{r}
#Run MANOVA
manova_panas <- manova(cbind(positive_mean, negative_mean) ~ cognition * conservation * species, data = data_summary)

#Extract Pillai's Trace and then build results table
pillai <- summary(manova_panas, test = "Pillai")$stats
pillai_df <- as.data.frame(pillai[rownames(pillai) != "Residuals", ])

manova_panas_df <- data.frame(
  Source       = rownames(pillai_df),
  Pillai       = round(pillai_df$Pillai, 3),
  F_approx     = round(pillai_df$`approx F`, 3),
  df1          = pillai_df$`num Df`,
  df2          = pillai_df$`den Df`,
  p_value      = round(pillai_df$`Pr(>F)`, 3)) %>%
  mutate(
    Eta2_partial = round((F_approx * df1) / (F_approx * df1 + df2), 3),
    p_value      = ifelse(p_value < 0.001, "<.001", as.character(p_value)))
manova_panas_df

print(manova_panas_df)
```


## Likert item analyses (ANOVAs + MANOVA)

Prepare Likert data for analysis
```{r}
#Convert Likert scale responses to numeric values
likert_conversion <- c("Strongly disagree" = 1, "Somewhat disagree" = 2, 
                       "Neither agree nor disagree" = 3, "Somewhat agree" = 4, 
                       "Strongly agree" = 5)

data_quant <- data_quant %>%
  mutate(across(l_1:l_18, ~ likert_conversion[.x])) #the columns containing likert responses 

#Descriptive statistics (mean and standard deviation for each likert item per condition)
likert_summary <- data_quant %>%
  group_by(message) %>%
  summarise(across(l_1:l_18,
                   list(mean = ~mean(.x, na.rm = TRUE), 
                        sd   = ~sd(.x, na.rm = TRUE))))

#Z-score the likert items (the z-scored likert items are now saved in a new dataframe called data_quant_z, which is what I call in the analysis)
likert_cols <- paste0("l_", 1:18)
data_quant_z <- data_quant %>%
  mutate(across(all_of(likert_cols), ~ as.numeric(scale(.x))))

#Make sure species is factors
data_quant_z$species <- as.factor(data_quant_z$species)
class(data_quant_z$species) #should say "factor"

#Create binary factors from message condition for 2x2 design (same as PANAS)
data_quant_z <- data_quant_z %>%
  mutate(
    cognition    = factor(ifelse(message %in% c("Cognition", "Combined"), "yes", "no"),
                          levels = c("no", "yes")),
    conservation = factor(ifelse(message %in% c("Conservation", "Combined"), "yes", "no"),
                          levels = c("no", "yes")))

table(data_quant_z$message, data_quant_z$cognition, data_quant_z$conservation) #verify each message levels maps to one cell
```


Check ANOVA assumptions
```{r}
##1. Multicollinearity (VIF scores)
#Fit model on just one outcome (same predictor variables for each)
vif_fit <- lm(l_1 ~ cognition * conservation * species,
              data = data_quant_z)
vif(vif_fit, type = 'predictor') #no multicollinearity because predictor variables from randomized video assignment

##2. Normality of residuals (Shapiro-Wilk Test) and Homogeneity of variance (Levenne's Test)
# Create a single grouping variable representing the 2x2 cells for levene's test
data_quant_z <- data_quant_z %>%
  mutate(group = interaction(cognition, conservation, species, drop = TRUE))

assumption_likert <- map_dfr(paste0("l_", 1:18), ~ {
  item <- .x
fit <- lm(
      as.formula(paste(item,
                       "~ cognition * conservation * species")),
      data = data_quant_z)
    # Shapiro-Wilk on residuals
    sw    <- shapiro.test(residuals(fit))
    # Levene's test
    lev <- leveneTest(data_quant_z[[item]], data_quant_z$group)
    
    data.frame(
      Item        = item,
      SW_W        = round(sw$statistic, 3),
      SW_p        = round(sw$p.value, 3),
      Levene_F    = round(lev$`F value`[1], 3),
      Levene_p    = round(lev$`Pr(>F)`[1], 4),
      stringsAsFactors = FALSE)})

print(assumption_likert)
#all items pass levene's test, and all items fail shapiro wilk test, which is expected given its sensitivity - using QQ plots to test normality of residuals instead (see below)

#Assessing normality of residuals with QQ plots
plots <- lapply(paste0("l_", 1:18), function(item) {
  fit <- lm(
    as.formula(paste(item, "~ cognition * conservation * species")),
    data = data_quant_z)
  res <- residuals(fit)
  ggplot(data.frame(res = res), aes(sample = res)) +
    stat_qq(size = 0.8) +
    stat_qq_line(color = "red") +
    ggtitle(item) +
    theme_minimal()
})
wrap_plots(plots, ncol = 6)
#all QQ plots show approximately normal distribution of residuals
```

Run Type III univariate ANOVAs on Likert items
```{r}
options(contrasts = c("contr.sum", "contr.poly"))

View(data_quant)

anova_likert <- map_dfr(likert_cols, ~ {
  question <- .x
  # Fit model
  fit <- lm(as.formula(paste(question, "~ cognition * conservation * species")),
            data = data_quant_z)
  
  # Type 3 ANOVA
  aov_results <- Anova(fit, type = 3)
  
  # Build ANOVA dataframe
  anova_df <- as.data.frame(aov_results)
  anova_df$Source <- rownames(anova_df)
  anova_df$MS <- anova_df$`Sum Sq` / anova_df$Df
 
  # Partial eta squared
  eta_df <- as.data.frame(eta_squared(aov_results, partial = TRUE, ci = NULL))
  eta_df$Source <- eta_df$Parameter

  # Format the final dataframe 
  final_df <- dplyr::left_join(anova_df, eta_df, by = "Source")
  final_df <- final_df[, c("Source", "Sum Sq", "Df", "MS", "F value", "Pr(>F)", "Eta2_partial")]
  final_df$Question <- question
  final_df <- final_df[, c("Question", "Source", "Sum Sq", "Df", "MS", "F value", "Pr(>F)", "Eta2_partial")]
  final_df <- final_df %>%
    mutate(across(where(is.numeric), ~ round(., 3)))
  final_df[is.na(final_df)] <- ""
  final_df
})
anova_likert
```

Run MANOVA on Likert items
```{r}
manova_likert <- manova(cbind(l_1, l_2, l_3, l_4, l_5, l_6, l_7, l_8, l_9, l_10, l_11, l_12, l_13, l_14, l_15, l_16, l_17, l_18)
                        ~ cognition * conservation * species, data = data_quant_z)

#Extract Pillai's Trace and then build results table
pillai <- summary(manova_likert, test = "Pillai")$stats
pillai_df <- as.data.frame(pillai[rownames(pillai) != "Residuals", ])

manova_likert_df <- data.frame(
  Source       = rownames(pillai_df),
  Pillai       = round(pillai_df$Pillai, 3),
  F_approx     = round(pillai_df$`approx F`, 3),
  df1          = pillai_df$`num Df`,
  df2          = pillai_df$`den Df`,
  p_value      = round(pillai_df$`Pr(>F)`, 3)) %>%
  mutate(
    Eta2_partial = round((F_approx * df1) / (F_approx * df1 + df2), 3),
    p_value      = ifelse(p_value < 0.001, "<.001", as.character(p_value)))
manova_likert_df
```


## Mediation models assessing message type -> positive/negative affect -> attitude

Create 3 attitude composites from Likert-scale items (traditional pro-conservation attitudes; cognition-influenced pro-conservation attitudes; empathy towards species)
```{r}
#Convert Likert scale responses to numeric values 
likert_conversion <- c("Strongly disagree" = 1, "Somewhat disagree" = 2, 
                       "Neither agree nor disagree" = 3, "Somewhat agree" = 4, 
                       "Strongly agree" = 5)

data_quant <- data_quant %>%
  mutate(across(l_1:l_18, ~ likert_conversion[.x]))

#Create attitude composites from Likert data (these items were theoretically selected based on item content)
data_quant <- data_quant %>%
  mutate(
    conservation_attitude_mean = rowMeans(select(., l_1, l_16), na.rm = TRUE),
    cognition_attitude_mean    = rowMeans(select(., l_7, l_8, l_9, l_18), na.rm = TRUE),
    empathy_mean                = rowMeans(select(., l_4, l_11, l_14, l_17), na.rm = TRUE)
  ) %>% #z-score composites
  mutate(
    conservation_attitude_z = as.numeric(scale(conservation_attitude_mean)),
    cognition_attitude_z    = as.numeric(scale(cognition_attitude_mean)),
    empathy_z                = as.numeric(scale(empathy_mean))
  )

#Cronbach's alpha for each composite
alpha_conservation <- alpha(select(data_quant, l_1, l_16))
alpha_cognition    <- alpha(select(data_quant, l_7, l_8, l_9, l_18))
alpha_empathy       <- alpha(select(data_quant, l_4, l_11, l_14, l_17))

cat("Conservation Attitude alpha:", round(alpha_conservation$total$raw_alpha, 3), "\n")
cat("Cognition-Influenced Attitude alpha:", round(alpha_cognition$total$raw_alpha, 3), "\n")
cat("Empathy alpha:", round(alpha_empathy$total$raw_alpha, 3), "\n")

# Inspect item-level drop-if-deleted stats — useful if any alpha is below ~.70
alpha_conservation$alpha.drop
alpha_cognition$alpha.drop
alpha_empathy$alpha.drop
```

Calculate Spearman-Brown split-half reliability coefficient for the two-item composite
```{r}
r_conservation <- cor(data_quant$l_1, data_quant$l_16, use = "complete.obs")
sb_conservation <- (2 * r_conservation) / (1 + r_conservation)
cat("Spearman-Brown reliability:", round(sb_conservation, 3), "\n")

# For comparison, also pull the alpha you already have
cat("Cronbach's alpha:", round(alpha_conservation$total$raw_alpha, 3), "\n")
```


Prepare and run mediation models
```{r}
#Recode message as a factor with Control as reference
data_quant_2 <- data_quant %>%
  filter(type == "final")

data_quant_2 <- data_quant_2 %>%
  mutate(message_f = factor(message, 
                            levels = c("Control", "Cognition", 
                                       "Conservation", "Combined")))
# Verify
table(data_quant_2$message_f)

#Add z-socred positive and negative PANAS means to this dataaset
data_quant_2 <- data_quant_2 %>% 
  mutate(
    positive_mean = rowMeans(select(., e_1, e_3, e_5, e_7, e_9), na.rm = TRUE),
    negative_mean = rowMeans(select(., e_2, e_4, e_6, e_8, e_10), na.rm = TRUE))

#z-score means
data_quant_2 <- data_quant_2 %>%
  mutate(
    positive_mean_z = as.numeric(scale(positive_mean)),
    negative_mean_z = as.numeric(scale(negative_mean)))

###Run three separate mediation models per mediator/outcome combination, each contrasting one active message against control (a total of 18 models)

#Define pathways
outcomes  <- c("conservation_attitude_z", "cognition_attitude_z", "empathy_z")
mediators <- c("positive_mean_z", "negative_mean_z")
treatments <- c("CognitionvsControl", "ConservationvsControl", "CombinedvsControl")

#2. Create dummy variables for each contrast against control
data_quant_2 <- data_quant_2 %>%
  mutate(
    CognitionvsControl    = as.integer(message_f == "Cognition"),
    ConservationvsControl = as.integer(message_f == "Conservation"),
    CombinedvsControl     = as.integer(message_f == "Combined"))

#3. Run mediation models and store results
mediation_summaries_v2 <- list()

set.seed(123)

for (outcome in outcomes) {
  for (mediator in mediators) {
    for (treat in treatments) {
      
      # Include all three dummies so each contrast is adjusted for the others
      other_treats <- setdiff(treatments, treat)
      
      mediator_formula <- as.formula(paste(
        mediator, "~", treat, "+", paste(other_treats, collapse = " + ")))
      
      outcome_formula <- as.formula(paste(
        outcome, "~", treat, "+", paste(other_treats, collapse = " + "), "+", mediator))
      
      mediator_model <- lm(mediator_formula, data = data_quant_2)
      outcome_model  <- lm(outcome_formula,  data = data_quant_2)
      
      med_result <- mediate(
        mediator_model,
        outcome_model,
        treat         = treat,
        mediator      = mediator,
        boot          = TRUE,
        sims          = 500, #*NOTE* when sims=5000 this takes a long time to run - the analyses in the paper used sims = 5000, but for the sake of this RMD file I'm putting sims=500 so it compiles more quickly
        control.value = 0,
        treat.value   = 1
      )
      
      key <- paste(treat, mediator, outcome, sep = "__")
      mediation_summaries_v2[[key]] <- summary(med_result)
      
      cat("\n=====================================\n")
      cat("Treatment:", treat, "| Mediator:", mediator, "| Outcome:", outcome, "\n")
      cat("=====================================\n")
      print(summary(med_result))
    }
  }
}

#Extract results table
results_table_v2 <- map_dfr(names(mediation_summaries_v2), function(key) {
  parts <- strsplit(key, "__")[[1]]
  extract_mediation_stats(mediation_summaries_v2[[key]], parts[1], parts[2], parts[3])
}) %>%
  mutate(ACME_p_adj = round(p.adjust(ACME_p, method = "BH"), 4))

print(results_table_v2)
```


## Petition signing analysis and visualization

Petition stacked bar plot
```{r}
#Count petition signage per message condition
data_quant %>%
  count(petition)

# Recode petition to simplified categories
data_petition_plot <- data_quant %>%
  mutate(
    petition_cat = case_when(
      str_detect(petition, "do not wish") ~ "No petition",
      str_detect(petition, "cognitive abilities of chimpanzees|cognitive abilities of the") ~ "Cognition",
      str_detect(petition, "to both protect") ~ "Combined",
      str_detect(petition, "protect chimpanzees|protect ʻAlalā") ~ "Conservation"
    ),
    petition_cat = factor(petition_cat, levels = c("Combined", "Conservation", "Cognition", "No petition")),
    message = factor(message, levels = c("Cognition", "Combined", "Conservation", "Control")),
    species = factor(species, levels = c("Chimpanzee", "ʻAlalā"))
  )

# Get counts
petition_plot_counts <- data_petition_plot %>%
  count(message, petition_cat) %>%
  group_by(message) %>%
  mutate(total = sum(n))

# Plot
petition_plot <-  ggplot(petition_plot_counts, aes(x = message, y = n, fill = petition_cat)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            color = "black", size = 5) +
  scale_fill_manual(
    values = c(
"No petition"  = "#999999",  # gray
  "Conservation" = "#0072B2",  # blue
  "Cognition"    = "#F0E442",  # yellow
  "Combined"     = "#D55E00"
    ),
    name = "Petition Choice"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Message", y = "Participant Count") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position    = "right",
    axis.text.x        = element_text(angle = 45, hjust = 1)
  )

petition_plot

ggsave("petition_plot.png", petition_plot, width = 200, height =150, units = "mm", dpi = 300)
```

Assess if there is an association between petition choice and study type 
```{r}
#Build contingency table and run chi square test
petition_studytype_contin <- table(data_petition_plot$type, data_petition_plot$petition_cat)
petition_studytype_contin

chisq.test(petition_studytype_contin)
#non-significant association, justifying the combination of petition choice data across study 1 and 2 
```

Assess if there is an association between petition choice and message condition, and between petition choice and species
```{r}
#Message condition
petition_message_contin <- table(data_petition_plot$message, data_petition_plot$petition_cat)
petition_message_contin

chisq.test(petition_message_contin)

#Species
petition_species_contin <- table(data_petition_plot$species, data_petition_plot$petition_cat)
petition_species_contin

chisq.test(petition_species_contin)

#non-significant association for both
```

## Donation choice analysis

Assess if there is an association between donation choice and study type
```{r}
##Combine donation choice categories into four groups (conservation, cognition, both, none) so donation choice is not split by species
data_quant <- data_quant %>%
  mutate(donation_group = case_when(
    donation_choice %in% c(
      "Chimpanzee conservation and research on chimpanzee cognition: donation website",
      "ʻAlalā conservation and research on ʻAlalā cognition: donation website"
    ) ~ "both",
    donation_choice %in% c(
      "Chimpanzee conservation: donation website",
      "ʻAlalā conservation: donation website"
    ) ~ "conservation",
    donation_choice %in% c(
      "Research on chimpanzee cognition: donation website",
      "Research on ʻAlalā cognition: donation website"
    ) ~ "cognition",
    donation_choice == "No I don't want to donate" ~ "none",
    TRUE ~ NA_character_
  ))

#turn into a factor 
data_quant$donation_group <- factor(
  data_quant$donation_group,
  levels = c("none", "conservation", "cognition", "both"))


##Build contingency table and run chi square test
donate_studytype_contin <- table(data_quant$type, data_quant$donation_group)
chisq.test(donate_studytype_contin)

#small cell sizes...confirm with fisher's exact test
fisher.test(donate_studytype_contin, simulate.p.value = TRUE, B = 10000) #p=0.7536
```
Non-significant association, okay to combine across study 1 and 2

# Qualitative Analysis

## Word association analysis + visualization 

Load word association data 
```{r}
#word association data from study 2
data_words2 <- read.csv("data_main/words_study2.csv",
                       header=TRUE,
                       stringsAsFactors=TRUE, 
                       sep=",",
                       na.strings="NA",
                       dec=".",
                       strip.white=TRUE)

#word association data from study 1
data_words1 <- read.csv("data_main/words_study1.csv",
                       header=TRUE,
                       stringsAsFactors=TRUE, 
                       sep=",",
                       na.strings="NA",
                       dec=".",
                       strip.white=TRUE)
```


Chi square test assessing association between study type and word sentiment
```{r}
#Combine word association data from study 1 and 2
data_words_combined <- bind_rows(data_words1, data_words2)

#Create "sentiment" column (contains positive, negative, or neutral)
data_words_combined <- data_words_combined %>%
  pivot_longer(
    cols = c(Positive, Negative, Neutral),
    names_to = "sentiment",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

#Build contingency table and run chi square test
wordsentiment_studytype_contin <- table(data_words_combined$type, data_words_combined$sentiment)
wordsentiment_studytype_contin

wordsentiment_studytype_chisq <- chisq.test(wordsentiment_studytype_contin)
wordsentiment_studytype_chisq

# Cramer's V
n <- sum(wordsentiment_studytype_contin)
sqrt(wordsentiment_studytype_chisq$statistic / (n * (2 - 1)))
```
Since effect size is 0.148 (nearing threshold of 0.15) we will analyze the word association data separately per study type


### Study 2 word association chi square tests + visualiation

Association btwn sentiment and message type
```{r}
#Create "sentiment" column (contains positive, negative, or neutral)
data_words2 <- data_words2 %>%
  pivot_longer(
    cols = c(Positive, Negative, Neutral),
    names_to = "sentiment",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

#Build contingency table and run chi square test
wordsentiment2_message_contin <- table(data_words2$message, data_words2$sentiment)
wordsentiment2_message_contin

wordsentiment2_message_test <- chisq.test(wordsentiment2_message_contin)
wordsentiment2_message_test

#View standardized residuals to assess directionality
wordsentiment2_message_test$stdres
```

Association btwn sentiment and species
```{r}
#Build contingency table and run chi square test
wordsentiment2_species_contin <- table(data_words2$species, data_words2$sentiment)
wordsentiment2_species_contin

wordsentiment2_species_test <- chisq.test(wordsentiment2_species_contin)
wordsentiment2_species_test

#View standardized residuals to assess directionality
wordsentiment2_species_test$stdres
```

Association btwn sentiment and species+message groups
```{r}
#create species+message groups
data_words2 <- data_words2 %>%
  mutate(species_message = paste(species, message, sep = "_"))

# check you actually have 8 groups
table(data_words2$species_message)

#Build contingency table and run chi square test
wordsentiment2_group_contin <- table(data_words2$species_message, data_words2$sentiment)
wordsentiment2_group_contin

wordsentiment2_group_test <- chisq.test(wordsentiment2_group_contin)
wordsentiment2_group_test

#View standardized residuals to assess directionality
wordsentiment2_group_test$stdres
```

Word association data visualization - study 2
```{r}
# Summarize the counts of each sentiment by group
wordsentiment2_summary <- data_words2 %>%
  group_by(species, message) %>%
  summarise(
    Positive = sum(Positive, na.rm = TRUE),
    Negative = sum(Negative, na.rm = TRUE),
    Neutral  = sum(Neutral, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Positive, Negative, Neutral),
               names_to = "sentiment",
               values_to = "count")

#Plot sentiment distribution of words (study 2)
wordsentiment2_prop <- wordsentiment2_summary %>%
  group_by(species, message) %>%
  mutate(prop = count / sum(count))

wordsentiment2_plot <- ggplot(wordsentiment2_prop,
                         aes(x = message, y = prop, fill = sentiment)) +
  geom_col() +
  facet_wrap(~ species) +
  labs(x = "Message", y = "Proportion", fill = "Sentiment") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c(
    Positive = "#4476a8",
    Negative = "#DDCC77",
    Neutral  = "#82cae9")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

wordsentiment2_plot

ggsave("wordsentiment2_plot.png", width = 200, height =150, units = "mm", dpi = 300)
```


### Study 1 word association chi square tests + visualiation

Association btwn sentiment and message type
```{r}
#Create "sentiment" column (contains positive, negative, or neutral)
data_words1 <- data_words1 %>%
  pivot_longer(
    cols = c(Positive, Negative, Neutral),
    names_to = "sentiment",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

View(data_words1)

#Build contingency table and run chi square test
wordsentiment1_message_contin <- table(data_words1$message, data_words1$sentiment)
wordsentiment1_message_contin

wordsentiment1_message_test <- chisq.test(wordsentiment1_message_contin)
wordsentiment1_message_test

#View standardized residuals to assess directionality
wordsentiment1_message_test$stdres
```

Association btwn sentiment and species
```{r}
#Build contingency table and run chi square test
wordsentiment1_species_contin <- table(data_words1$species, data_words1$sentiment)
wordsentiment1_species_contin

wordsentiment1_species_test <- chisq.test(wordsentiment1_species_contin)
wordsentiment1_species_test

#View standardized residuals to assess directionality
wordsentiment1_species_test$stdres
```

Association btwn sentiment and species+message groups
```{r}
#create species+message groups
data_words1 <- data_words1 %>%
  mutate(species_message = paste(species, message, sep = "_"))

# check you actually have 8 groups
table(data_words1$species_message)

#Build contingency table and run chi square test
wordsentiment1_group_contin <- table(data_words1$species_message, data_words1$sentiment)
wordsentiment1_group_contin

wordsentiment1_group_test <- chisq.test(wordsentiment1_group_contin)
wordsentiment1_group_test

#View standardized residuals to assess directionality
wordsentiment1_group_test$stdres
```

Word association data summary + visualization - study 1
```{r}
# Summarize the counts of each sentiment by group
wordsentiment1_summary <- data_words1 %>%
  group_by(species, message) %>%
  summarise(
    Positive = sum(Positive, na.rm = TRUE),
    Negative = sum(Negative, na.rm = TRUE),
    Neutral  = sum(Neutral, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Positive, Negative, Neutral),
               names_to = "sentiment",
               values_to = "count")

#Plot sentiment distribution of words (study 1)
wordsentiment1_prop <- wordsentiment1_summary %>%
  group_by(species, message) %>%
  mutate(prop = count / sum(count))

wordsentiment1_plot <- ggplot(wordsentiment1_prop,
                         aes(x = message, y = prop, fill = sentiment)) +
  geom_col() +
  facet_wrap(~ species) +
  labs(x = "Message", y = "Proportion", fill = "Sentiment") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c(
    Positive = "#4476a8",
    Negative = "#DDCC77",
    Neutral  = "#82cae9")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

wordsentiment1_plot

ggsave("wordsentiment1_plot.png", width = 200, height =150, units = "mm", dpi = 300)
```

## Video short response analysis + visualization

Load short response about video data
```{r}
#short response about video data from study 2
data_sr2 <- read.csv("data_main/videoSR_study2.csv",
                       header=TRUE,
                       stringsAsFactors=TRUE, 
                       sep=",",
                       na.strings="NA",
                       dec=".",
                       strip.white=TRUE)

#short response about video data from study 1
data_sr1 <- read.csv("data_main/videoSR_study1.csv",
                       header=TRUE,
                       stringsAsFactors=TRUE, 
                       sep=",",
                       na.strings="NA",
                       dec=".",
                       strip.white=TRUE)
```

### Video short response sentiment analyses 

Chi square test assessing association between study type and word sentiment
```{r}
#Combine short response data from study 1 and 2
data_sr_combined <- bind_rows(data_sr1, data_sr2)

#Create "sentiment" column (contains positive, negative, or neutral)
data_sr_combined <- data_sr_combined %>%
  pivot_longer(
    cols = c(Positive, Negative, Neutral),
    names_to = "sentiment",
    values_to = "value"
  ) %>%
  filter(!is.na(value))

#Build contingency table and run chi square test
srsentiment_studytype_contin <- table(data_sr_combined$type, data_sr_combined$sentiment)
srsentiment_studytype_contin 

srsentiment_studytype_chisq <- chisq.test(srsentiment_studytype_contin)
srsentiment_studytype_chisq

# Cramer's V
n <- sum(srsentiment_studytype_contin)
sqrt(srsentiment_studytype_chisq$statistic / (n * (2 - 1)))
```

We are combining short response data across study 1 and 2 due to small effect size, and as such will be conducing the chi square tests on the data_sr_combined dataset

Association btwn sentiment and message type
```{r}
#Build contingency table and run chi square test
srsentiment_message_contin <- table(data_sr_combined$Message, data_sr_combined$sentiment)
srsentiment_message_contin

srsentiment_message_test <- chisq.test(srsentiment_message_contin)
srsentiment_message_test 

#View standardized residuals to assess directionality
srsentiment_message_test$stdres
```

Association btwn sentiment and species
```{r}
#Build contingency table and run chi square test
srsentiment_species_contin <- table(data_sr_combined$species, data_sr_combined$sentiment)
srsentiment_species_contin

srsentiment_species_test <- chisq.test(srsentiment_species_contin)
srsentiment_species_test

#View standardized residuals to assess directionality
srsentiment_species_test$stdres
```

Association btwn sentiment and species+message groups
```{r}
#create species+message groups
data_sr_combined <- data_sr_combined %>%
  mutate(species_message = paste(species, Message, sep = "_"))

# check you actually have 8 groups
table(data_sr_combined$species_message)

#Build contingency table and run chi square test
srsentiment_group_contin <- table(data_sr_combined$species_message, data_sr_combined$sentiment)
srsentiment_group_contin

srsentiment_group_test <- chisq.test(srsentiment_group_contin)
srsentiment_group_test

#View standardized residuals to assess directionality
srsentiment_group_test$stdres
```

Short response sentiment data visualization
```{r}
#Plot sentiment distribution of words (study 2)
srsentiment_prop <- data_sr_combined %>%
  count(species, Message, sentiment, name = "n") %>%
  group_by(species, Message) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

srsentiment_plot <- ggplot(srsentiment_prop,
                         aes(x = Message, y = prop, fill = sentiment)) +
  geom_col() +
  facet_wrap(~ species) +
  labs(x = "Message", y = "Proportion", fill = "Sentiment") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c(
    Positive = "#4476a8",
    Negative = "#DDCC77",
    Neutral  = "#82cae9")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

srsentiment_plot

ggsave("wordsentiment2_plot.png", width = 200, height =150, units = "mm", dpi = 300)
```

### Video short response axial code analyses

Axial code chi square analysis for short response about video
```{r}
#Create "code" column (contains the five axial code columns)
data_sr_combined <- data_sr_combined %>%
  mutate(code = case_when(
    Informational == 1 ~ "Informational",
    Interest == 1 ~ "Interest",
    Empathy == 1 ~ "Empathy",
    Action == 1 ~ "Action",
    Disengaged == 1 ~ "Disengaged",
    TRUE ~ NA_character_
  ))

#verify:
#row count should be unchanged from before this step
nrow(data_sr_combined)

# each row's five code columns should sum to exactly 1
table(rowSums(data_sr_combined[, c("Informational","Interest","Empathy","Action","Disengaged")], na.rm = TRUE))
```

Association btwn axial code and message type
```{r}
#Build contingency table and run chi square test
srcode_message_contin <- table(data_sr_combined$Message, data_sr_combined$code)
srcode_message_contin

srcode_message_test <- chisq.test(srcode_message_contin)
srcode_message_test

#View standardized residuals to assess directionality
srcode_message_test$stdres

#small cell sizes...two options to confirm results: 
#confirm with fisher's exact test
fisher.test(srcode_message_contin, simulate.p.value = TRUE, B = 10000)

#alternatively, simulate p - value
chisq.test(srcode_message_contin, simulate.p.value = TRUE, B = 10000)
```

Association btwn axial code and species
```{r}
#Build contingency table and run chi square test
srcode_species_contin <- table(data_sr_combined$species, data_sr_combined$code)
srcode_species_contin

srcode_species_test <- chisq.test(srcode_species_contin)
srcode_species_test

#View standardized residuals to assess directionality
srcode_species_test$stdres
```

Association btwn axial code and species+message groups
```{r}
#create species+message groups
data_sr_combined <- data_sr_combined %>%
  mutate(species_message = paste(species, Message, sep = "_"))

# check you actually have 8 groups
table(data_sr_combined$species_message)

#Build contingency table and run chi square test
srcode_group_contin <- table(data_sr_combined$species_message, data_sr_combined$code)
srcode_group_contin

srcode_group_test <- chisq.test(srcode_group_contin)
srcode_group_test

#View standardized residuals to assess directionality
srcode_group_test$stdres

#small cell sizes...two options to confirm results: 
#confirm with fisher's exact test
fisher.test(srcode_group_contin, simulate.p.value = TRUE, B = 10000)

#alternatively, simulate p - value
chisq.test(srcode_group_contin, simulate.p.value = TRUE, B = 10000)
```

Stacked bar plot showing proportion of axial codes per message type faceted by species
```{r}
srcode_prop <- data_sr_combined %>%
  filter(!is.na(code)) %>%
  count(species, Message, code, name = "n") %>%
  group_by(species, Message) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

srcode_plot <- ggplot(srcode_prop, aes(x = Message, y = prop, fill = code)) +
  geom_col() +
  facet_wrap(~ species) +
  labs(x = "Message", y = "Proportion", fill = "Theme") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c(
    "Informational" = "#E69F00",   
    "Interest" = "#CC79A7",  
    "Empathy" = "#009E73", 
    "Action" = "#D55E00",        
    "Disengaged" = "#56B4E9")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

srcode_plot
ggsave("srcode_plot.png", width = 200, height =150, units = "mm", dpi = 300)
```

## Donation short response axial code analyses

Load donation short response data
```{r}
#donation short response data (combined study 1 and 2)
data_donation <- read.csv("data_main/donationSR_combined.csv",
                       header=TRUE,
                       stringsAsFactors=TRUE, 
                       sep=",",
                       na.strings="NA",
                       dec=".",
                       strip.white=TRUE)
```

Axial code chi square analysis for short response about video
```{r}
#Create "code" column (contains the five axial code columns)
data_donation <- data_donation %>%
  mutate(code = case_when(
    Alternate.Support == 1 ~ "Alternate.Support",
    Financial.Constraints == 1 ~ "Financial.Constraints",
    Format.Critique == 1 ~ "Format.Critique",
    Information.Gap == 1 ~ "Information.Gap",
    Low.Motivation == 1 ~ "Low.Motivation",
    TRUE ~ NA_character_
  ))

#verify:
#row count should be unchanged from before this step
nrow(data_donation)
```

Association between axial code and study type (justificaiton for combining study 1 and 2 responses)
```{r}
#Build contingency table and run chi square test
donatecode_study_contin <- table(data_donation$type, data_donation$code)
chisq.test(donatecode_study_contin)
```

Association btwn axial code and message type
```{r}
#Build contingency table and run chi square test
donatecode_message_contin <- table(data_donation$message, data_donation$code)
donatecode_message_contin

chisq.test(donatecode_message_contin)

#small cell sizes...two options to confirm results: 
#confirm with fisher's exact test
fisher.test(donatecode_message_contin, simulate.p.value = TRUE, B = 10000)

#alternatively, simulate p - value
chisq.test(donatecode_message_contin, simulate.p.value = TRUE, B = 10000)
```

Association btwn axial code and species
```{r}
#Build contingency table and run chi square test
donatecode_species_contin <- table(data_donation$species, data_donation$code)
donatecode_species_contin

chisq.test(donatecode_species_contin)
```

Association btwn axial code and species+message groups
```{r}
#create species+message groups
data_donation<- data_donation %>%
  mutate(species_message = paste(species, message, sep = "_"))

# check you actually have 8 groups
table(data_donation$species_message)

#Build contingency table and run chi square test
donatecode_group_contin <- table(data_donation$species_message, data_donation$code)
donatecode_group_contin

chisq.test(donatecode_group_contin)

#small cell sizes...two options to confirm results: 
#confirm with fisher's exact test
fisher.test(donatecode_group_contin, simulate.p.value = TRUE, B = 10000)

#alternatively, simulate p - value
chisq.test(donatecode_group_contin, simulate.p.value = TRUE, B = 10000)
```

Stacked bar plot visualization of donation explanation short response axial codes
```{r}
data_donation_plot <- data_donation %>%
  pivot_longer(
    cols = c(Financial.Constraints,
      Information.Gap,
      Low.Motivation,
      Format.Critique,
      Alternate.Support),
    names_to = "code_plot",
    values_to = "value"
  ) %>%
  mutate(code = gsub("\\.", " ", code)) %>% #adds space between legend items
  filter(!is.na(value))

#create proportion of donation themes per participant for plotting
data_donation_prop <- data_donation_plot %>%
  group_by(species, message, code_plot) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(species, message) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

#plot
donation_themes <- ggplot(data_donation_prop,
       aes(x = message, y = prop, fill = code_plot)) +
  geom_col() +
  facet_wrap(~ species) +
  labs(
    x = "Message",
    y = "Proportion",
    fill = "Theme"
  ) +
  theme_minimal(base_size = 14) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_fill_manual(values = c(
    "Alternate.Support" = "#E69F00",
    "Financial.Constraints" = "#CC79A7",
    "Format.Critique" = "#009E73",
    "Information.Gap" = "#D55E00",
    "Low.Motivation" = "#56B4E9"
  )) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
donation_themes

ggsave("donation_themes.png", width = 200, height =150, units = "mm", dpi = 300)
```

## Pilot survey power analysis

Load pilot survey data
```{r}
data_pilot <- read.csv("data/pilot/quant_chimpalala_combined.csv",
                       header=TRUE,
                       stringsAsFactors=TRUE, 
                       sep=",",
                       na.strings="NA",
                       dec=".",
                       strip.white=TRUE)

```

ANOVA to determine effect size
```{r}
#ANOVA to determine effect size for each likert question (compared across 4 message types)
likert_cols <- paste0("l_", 1:20)
anova_likert_message_eta <- map_dfr(likert_cols, function(var) {
  model <- aov(reformulate("message", var), data = data_pilot)
  anova_result <- summary(model)[[1]]
  
  # Compute partial eta squared 
  ss_between <- anova_result["message", "Sum Sq"]
  ss_within <- anova_result["Residuals", "Sum Sq"]
  eta_sq_partial <- ss_between / (ss_between + ss_within)
  
  data.frame(
    Question = var,
    p_value = anova_result["message", "Pr(>F)"],
    eta_sq = eta_sq_partial
  )
})
print(anova_likert_message_eta) #shows p-value and eta squared for each likert question 

#estimate average effect size (average eta)
mean_eta <- mean(anova_likert_message_eta$eta_sq, na.rm = TRUE)
f <- sqrt(mean_eta / (1 - mean_eta))
f 

#run ANOVA power analysis using calculated effect size to determine sample size
pwr.anova.test(k = 4, f = 0.2, sig.level = 0.05, power = 0.8)
```
We need ~89 participants per message type (~356 total) to achieved the desired power